---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

I will prepare the environment by loading the required libraries and sourcing the data assuming the activity.zip file is present in the working directory.  I will then cast the date column as a POSIX time, and remove the rows with 0 steps.

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}

#load required libraries
library(ggplot2)
library(plyr)
library(sqldf)

activity <- read.csv(unz("activity.zip", "activity.csv"), header =TRUE)

##need as a date to try stuff out
activity$date <- (as.POSIXct(as.character(activity$date), format = "%Y-%m-%d", tz = ""))

```


## What is mean total number of steps taken per day?

Plot a histogram to show the mean number of steps taken per day and then calculate the mean:

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
##histogram for total number of steps
ggplot(activity, aes(x=date, y=steps)) + geom_bar(stat="identity", color = "black")  + ggtitle("Total Number of steps over the time period")

##get the total number of steps per day

by_day <- sqldf("select sum(steps) as sum, date from activity group by date")

##and fix the column "sum" to be an integer
by_day$sum <- as.integer(by_day$sum)

##take the mean and the median

first_mean <- mean(by_day$sum, na.rm=TRUE)
first_median <- median(by_day$sum, na.rm=TRUE)

```

The mean/median number of total number of steps per day is therefore

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}

first_mean
first_median

```


## What is the average daily activity pattern?

I will aggregate the average number of steps per interval, and put this into a new dataframe.  I will then plot using the now familiar *ggplot* library.

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}

daily_pattern <- sqldf("select avg(steps) as average, interval from activity group by interval")

ggplot(daily_pattern, aes(x=interval, y=average)) + geom_bar(stat="identity", color = "black")  + ggtitle("Average number of steps per interval")

```

We can see there is a clear pattern for night and day, with peaks of activities around breakfast, lunch, mid-afternoon and finally early evening.

## Imputing missing values

First find out the number of rows with NA's:

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}

colSums(is.na(by_day))

```

My imputation strategy is to input the average of the total of daily steps over the observation period into those days that do not have a value:

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
##replace missing values for Mean_steps with the average

by_day$sum[is.na(by_day$sum)] <- round(first_mean, 0)

##new histogram for total number of steps

ggplot(by_day, aes(x=date, y=sum)) + geom_bar(stat="identity", color = "black")  + ggtitle("Total Number of steps over the time period after imputation")

##report on Mean and median

second_mean <- mean(by_day$sum, na.rm=TRUE)
second_median <- median(by_day$sum, na.rm=TRUE)

##report on difference

abs(second_mean - first_mean)
abs(second_median - first_median)

```

## Are there differences in activity patterns between weekdays and weekends?

```{r}

##add a column with the weekday
by_day$weekdays <- strtrim(weekdays(by_day$date), 3)

by_weekday <- sqldf("select avg(sum) as sum, weekdays from by_day group by weekdays")

## factor and order
by_weekday$weekdays <- factor(by_weekday$weekdays, levels= c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))

by_weekday <- by_weekday[order(by_weekday$weekdays), ]

##add a column to show if it's a weekend or not
by_weekday$weekend <- c("Workday", "Workday", "Workday", "Workday", "Workday", "Weekend", "Weekend")

##plot using facets
ggplot(by_weekday, aes(x=weekdays, y=sum)) + geom_bar(stat="identity", color = "blue", fill ="blue")  + ggtitle("Total Number of steps over the time period after imputation, by week and weekend") + facet_grid(. ~ weekend)

```

Clearly, the weekend is busier than the average office-based workday.

